---
title: "Final Project"
author: "Haodi Tu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '3'

---

# Research Question 

How can we prevent suicides?

# Clear environment

```{r}
# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)
```

# Load Packages

```{r echo = FALSE}
library(tidyverse)
library(stringr)
library(kableExtra)
library(GGally)
library(psych)
library(gridExtra)
library(janitor)
library(MultinomialCI)
library(htmlTable)
library(countrycode)
library(rworldmap)
library(sjPlot)
library(directlabels)
```

# Load data
```{r}
# Load data 
master <- read.csv("master.csv")


# Examine data structure
str(master)
```

# Things need to be done before EDA

+ Remane column names 
+ Change data type for some variables, like year
+ Data selection: 
    - recalculate suicide rate and replace column suicides.100k.pop
    - remove all missing data from HDI.for.year
    - Keep data that has 12 rows (2 sex * 6 age group) for every country.year
    - Eliminate coutries with less than 3 years of data
    - Adding continent for convenient purpose


```{r}
#rename columns
names(master) <- c('country','year','sex','age','suicides_no','population','suicide_rate',
                   'country_year', 'HDI_for_year','gdp_for_year','gdp_per_capita', 'generation')
# Convert column year data type to factor
#master['year'] = as.factor(master$year)
# Convert column gdp_for_year to numeric data type
master$gdp_for_year <- gsub(",","", master$gdp_for_year)  
master$gdp_for_year <-  as.numeric(master$gdp_for_year)

# Delete years in age values
master$age <- gsub(" years", "", master$age)
master$age <- factor(master$age,
                     ordered = T,
                     levels = c("5-14", "15-24","25-34", "35-54", "55-74", "75+"))

```

```{r}
# data selection 
# remove all missing data from HDI_for_year
sum(is.na(master$HDI_for_year)) # NA values = 19456 > 2/3 of total observations, so not applicable

# remove this column
master <- subset(master, select = -c(HDI_for_year, suicide_rate))

master$suicide_rate <- (master$suicides_no/master$population) *100

# keep data that has 12 rows (2 sex * 6 age group) for each country_year
master %>% 
  group_by(country_year) %>% 
  count() %>% 
  filter(n != 12) # Observations from year 2016 have only 10 rows, so fliter out 2016

master <- master %>% 
  filter(year != 2016) %>% 
  select(-country_year)

# Eliminate coutries with less than 3 years of data
missing_data_year <-master %>% 
  group_by(country) %>% 
  summarize(rows = n(),
            years = rows/ 12) %>% 
  arrange(years)

master <- master %>% 
  filter(!(country %in% head(missing_data_year$country,7)))

```

```{r}
# Adding continent
master$continent <- countrycode(sourcevar = master[, "country"],
                                origin = "country.name",
                                destination = "continent")  
master$continent <- as.factor(master$continent)

```

```{r}
master <- master %>%  
  mutate_if(is.integer, as.numeric)

str(master)

summary(master)

# Save csv file for power bi 
write.csv(master, "master_bi.csv")

```

Observations

+ Six factor variables, three integer variables and two numeric variables
+ Sample is drawed from 101 countries from 1985 to 2015, observations are equalily stored by gender
+ Number of observations from each age group are equaliy distributed, except for the youngest group (5-14 years) 
+ Suicide number has large range from 0 to 22338, left skewed 
+ Population has a large range vary by countries, left skewed
+ Suicide_ppl_100k is calculated from suicides number divided by population in 100,000 unit, left skewed
+ Removed Human Development Index because it has lots of NA's
+ Distributions of gdp_for_year and gdp_per_capita has similar distributions 
+ Five genderations, where Silent and Generation X has the highest observations
+ Five continents, Europe has the most observations and Africa has the least

    
Questions to be answered 

+ To find out possible suicide reasons
    
    + Suicide number across world
    + Relationship between suicide rate and variables (e.g. age, gdp per capita, gender)

    


# Univariable non-graphical - Categorical

Categorical data - tabulation

## Country 

```{r}
# Country table 
country_all <- master %>% 
  group_by(country) %>% 
  summarise(count = n(),
            tot_suicides = sum(suicides_no),
            avg_suicides_ppl = mean(suicide_rate),
            avg_gdp_year_100k= mean(gdp_for_year/100000),
            avg_gdp_ppl = mean(gdp_per_capita))
country_all
```

Observation

+ surprisingly, Dominica has 0 suicide record
+ Table is hard to see pattern with large country rows

## Year

```{r}
# Year table
year_all <- master %>% 
  group_by(year) %>% 
  summarise(count = n(),
            tot_suicides = sum(suicides_no),
            tot_ppl = sum(population),
            avg_suicides_ppl = mean(suicide_rate),
            avg_gdp_year_100k= mean(gdp_for_year/100000),
            avg_gdp_ppl = mean(gdp_per_capita))
            
year_all
```

Observation

+ Looks like suicide number has a bell shape, increase from 1985 and decease after 2000 year

## Sex

```{r}
# Sex table 
sex_all <- master %>% 
  group_by(sex) %>% 
  summarise(count = n(),
            tot_suicides = sum(suicides_no),
            avg_suicides_ppl = mean(suicide_rate))
            
sex_all

```

Observation

+ Suicides number of male is 4 times as suicides number of female

## Age

```{r}
# Age table
age_all <- master %>% 
  group_by(age) %>% 
  summarise(count = n(),
            tot_suicides = sum(suicides_no),
            avg_suicides_ppl = mean(suicide_rate))
            
age_all


```

Observation

+ Age group 35-54 years has the most suicide number
+ Age group 5-14 years has the lest suicide number
+ However, age group 55-74 years has the highest suidcide rate  


## Generation

```{r}
# Year table
generation_all <- master %>% 
  group_by(generation) %>% 
  summarise(count = n(),
            tot_suicides = sum(suicides_no),
            avg_suicides_ppl = mean(suicide_rate),
            avg_gdp_year_100k= mean(gdp_for_year/100000),
            avg_gdp_ppl = mean(gdp_per_capita))
            
generation_all
```

Observation

+ Boomers have the highest suicide number and Generation Z has the least
+ G.I Generation has the highest suicide rate

## Continent 

```{r}
# Continents table 
continent_all <- master %>% 
  group_by(continent) %>% 
  summarise(count = n(),
            tot_suicides = sum(suicides_no),
            avg_suicides_ppl = mean(suicide_rate),
            avg_gdp_year_100k= mean(gdp_for_year/100000),
            avg_gdp_ppl = mean(gdp_per_capita))
continent_all
```

Observation

+ Europe has the most suicide number and average suicide rate
+ Africa has the least

# Univariate graphical - Categorical 

## Country

```{r}
# Code histograms using grid.arrange so can see all quant variables together group by country

country_all %>% 
  ggplot(aes(x = country, y = tot_suicides)) +
    geom_bar(stat = "identity") +
  coord_flip()

```


Observations

+ still hard to see pattern
+ certain countries have very large suicide number
+ will conduct a global map or continent heat map with cuiside number to see patterns

## Year

```{r}
# Code histograms using grid.arrange so can see all quant variables together group by year
grid.arrange(
  year_all %>% 
    ggplot(aes(x = year, y = count)) +  
      geom_bar(stat = "identity"),
  year_all %>% 
    ggplot(aes(x = year, y = tot_suicides)) +
      geom_bar(stat = "identity"),
  year_all %>% 
    ggplot(aes(x = year, y = tot_ppl/1000000)) +
      geom_bar(stat = "identity"),
  year_all %>% 
    ggplot(aes(x = year, y = avg_suicides_ppl)) +
      geom_bar(stat = "identity"),
  year_all %>% 
    ggplot(aes(x = year, y = avg_gdp_year_100k)) +
      geom_bar(stat = "identity"),
  year_all %>% 
    ggplot(aes(x = year, y = avg_gdp_ppl)) +
      geom_bar(stat = "identity"),
  ncol = 2
)
```

Observations 

+ Average gdp grows over year, which is practical to real world
+ Total suicide number grow over the first 15 years, and slicely descrese after then 

## Sex

```{r}
# Code histograms using grid.arrange so can see all quant variables together group by Sex
grid.arrange(
  sex_all %>% 
    ggplot(aes(x = sex, y = count)) +  
      geom_bar(stat = "identity"),
  sex_all %>% 
    ggplot(aes(x = sex, y = tot_suicides)) +
      geom_bar(stat = "identity"),
  sex_all %>% 
    ggplot(aes(x = sex, y = avg_suicides_ppl)) +
      geom_bar(stat = "identity"),

  ncol = 2
)
```

Observations 

+ Male suicide number and rate is 5 times higher than female

## Age

```{r}
# Code histograms using grid.arrange so can see all quant variables together group by age
grid.arrange(
  age_all %>% 
    ggplot(aes(x = age, y = count)) +  
      geom_bar(stat = "identity"),
  age_all %>% 
    ggplot(aes(x = age, y = tot_suicides)) +
      geom_bar(stat = "identity"),
  age_all %>% 
    ggplot(aes(x = age, y = avg_suicides_ppl)) +
      geom_bar(stat = "identity"),
  ncol = 2
)
```

Observations 

+ Observations from all age group are same
+ The youngest age (5-14 years) group has the least suicide number and suicide rate
+ The oldest age (75+ years) group has the least suicide number but the highest suicide rate

## Generation

```{r}
# Code histograms using grid.arrange so can see all quant variables together group by generation
grid.arrange(
  generation_all %>% 
    ggplot(aes(x = generation, y = count)) +  
      geom_bar(stat = "identity") + coord_flip(),
  generation_all %>% 
    ggplot(aes(x = generation, y = tot_suicides)) +
      geom_bar(stat = "identity")+ coord_flip(),
  generation_all %>% 
    ggplot(aes(x = generation, y = avg_suicides_ppl)) +
      geom_bar(stat = "identity")+ coord_flip(),
  generation_all %>% 
    ggplot(aes(x = generation, y = avg_gdp_year_100k)) +
      geom_bar(stat = "identity")+ coord_flip(),
  generation_all %>% 
    ggplot(aes(x = generation, y = avg_gdp_ppl)) +
      geom_bar(stat = "identity")+ coord_flip(),
  ncol = 2
)
```

Observations 

+ small suicide number and rate on Generation z, maybe because of the growth of gdp
+ Boomers has large suicide number
+ G.I. generation has the highest average suicide rate under the lowest gdp

## Continent

```{r}
# Code histograms using grid.arrange so can see all quant variables together group by continent
grid.arrange(
  continent_all %>% 
    ggplot(aes(x = continent, y = count)) +  
      geom_bar(stat = "identity") + coord_flip(),
  continent_all %>% 
    ggplot(aes(x = continent, y = tot_suicides)) +
      geom_bar(stat = "identity")+ coord_flip(),
  continent_all %>% 
    ggplot(aes(x = continent, y = avg_suicides_ppl)) +
      geom_bar(stat = "identity")+ coord_flip(),
  continent_all %>% 
    ggplot(aes(x = continent, y = avg_gdp_year_100k)) +
      geom_bar(stat = "identity")+ coord_flip(),
  continent_all %>% 
    ggplot(aes(x = continent, y = avg_gdp_ppl)) +
      geom_bar(stat = "identity")+ coord_flip(),
  ncol = 2
)
```

Observations

+ Americas has the highest national gdp
+ Europe has the largest suicide body as well as suicide rate
+ Europe has the highest GDP per capita


# Univariate Non-graphical - Quantitative
```{r}
master %>% 
  select(-c(country, year, sex, age, generation)) %>% 
  summary()
```

Observation

+ Suicides_no - a large range, right skewed distribution, exist extreme numbers
+ Population - right skewed distribution 
+ suicide_rate - a large range, right skewed distribution 
+ gdp_for_year - a large range, right skewed distribution
+ gdp_per_capita - a large range, right skewed distribution



# Univariate Graphical - Quantitative 
```{r}
master %>% 
  ggplot(aes(suicides_no)) +
    geom_histogram(binwidth = 500)
```


```{r}
master %>% 
  ggplot(aes(population)) +
    geom_histogram(binwidth = 1000000)
```


```{r}
master %>% 
  ggplot(aes(suicide_rate)) +
    geom_histogram(binwidth = .001)
```


```{r}
master %>% 
  ggplot(aes(gdp_for_year)) +
    geom_histogram(binwidth = 100000000000)
```

```{r}
master %>% 
  ggplot(aes(gdp_per_capita)) +
    geom_histogram(binwidth = 100)
```


## Boxplots of quantitative data
```{r}
# Set grid pattern for graph arrangement
par(mfrow = c(2, 3))
boxplot(master$suicides_no, main = "suicide number")
boxplot(master$population, main = "population")
boxplot(master$suicide_rate, main = "suicide rate in 100k")
boxplot(master$gdp_for_year, main = "National Yearly GDP")
boxplot(master$gdp_per_capita, main = "Average GDP Per captia")
```

Observation 

+ Many outliers across quantitative variables, because situation differs by countries 

# Multivariate non-graphical -- Categorical

## Proportion Cross tabs


```{r}
addmargins(round(prop.table(xtabs(~ continent + year, data = master)), 4))

```


```{r}
addmargins(round(prop.table(xtabs(~ continent + sex, data = master)), 3))

```


```{r}
addmargins(round(prop.table(xtabs(~ continent + age, data = master)), 3))

```


```{r}
addmargins(round(prop.table(xtabs(~ age + sex, data = master)), 4))

```


# Multivariate graphical -- Categorical

```{r}
master %>%
  group_by(continent, year) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(continent, year)) + 
  geom_tile(aes(fill = -count))

```


```{r}
master %>%
  group_by(year, sex) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(year, sex)) + 
  geom_tile(aes(fill = -count))

```


# Multivariate non-graphical -- Quantitative


```{r}
# Select all numeric variables
sub_master <- select(master, "year","suicides_no", "population", "suicide_rate", "gdp_for_year", "gdp_per_capita")

cor(sub_master)
```



# Multivariate graphical code -- Quantitative 

```{r}
sub_master %>% 
  pairs.panels()
```

Observation

+ high correlated variables are suicides_no and population, population and gdp_for_year
+ Surprisingly, population and suicide_rate do not have high correlation, neither do suicide_rate and gdp_per_capita
+ GDP per captia increases as time goes by


```{r}
# Standard scatter plot

master %>% 
  ggplot(aes(x = population, y = suicide_rate, color = continent)) +
  geom_point() +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE)


```


Observation 

+ European countries have high suicide rates, regardless the population
+ Suicide rate grows slightly as population increase in Americas 
+ Africa has one dot at over 20% suicide rate
+ Asia has lower suicide rate than Europe in general

```{r}
p3 <- master %>% 
  group_by(continent) %>% 
  ggplot(aes(x = gdp_per_capita/1000, y = suicides_no, col = continent))+ 
  geom_smooth(method = "lm", se= FALSE)  + 
  geom_smooth(method = "lm",  aes(group = 1, colour = "Overall Slope"),se= FALSE) +
  ggtitle( "Relationship Between GDP Per Capita (in $1k) and Suicides",
       subtitle = "Overall positive slope, richer continents have more suicides") +
  xlab("GDP per capita in $1k") +
  ylab("Average Suicides") +
  theme_classic()

p3
ggsave('slope.png', p3)
```

Do richer countries have lower suicide rate?

+ Suiside rates goes down as GDP per capita increase, especially European and Africa countries
+ Continents, like Americas and Oceania, have a positive relationship between suicide rate and GDP 
+ Overall, a light positive relationship, indicating richer coutries have a higher suicide rate


# Multivariate non-graphical -- Categorical and Quantitative


```{r}
tot_suicide_no_continent_sex <- master %>%
  group_by(continent, sex) %>%
  summarise(tot_suicide_no = sum(suicides_no)) %>%
  spread(continent, tot_suicide_no) %>%
  as.data.frame()

tot_suicide_no_continent_sex
```



```{r}
tot_suicide_no_continent_age <- master %>%
  group_by(continent, age) %>%
  summarise(tot_suicide_no = sum(suicides_no)) %>%
  spread(continent, tot_suicide_no) %>%
  as.data.frame()

tot_suicide_no_continent_age
```




# Multivariate graphical -- Categorical and Quantitative


```{r}
master %>% 
  ggplot(aes(x = generation , y = suicides_no , fill = generation))+
  geom_boxplot(alpha = .50)+
  coord_cartesian(ylim = c(0,500)) +
  theme_bw()

```

## Distribution of suicide rates across age groups
```{r}
master %>% 
ggplot(aes(x = suicide_rate, fill = age, alpha = 0.5)) + 
geom_histogram(binwidth = 0.01) + 
facet_wrap(~age) +
guides(alpha = FALSE) +
labs(title = "Distribution of suicide rates across age groups", subtitle = "Lots of countries having low suicide rates for the youngest age group", x = "Suicide rates (in %)")

```

## Distribution of suicide rates across generation groups
```{r}

master %>% 
ggplot(aes(x = suicide_rate, fill = generation, alpha = 0.5)) + 
geom_histogram(position = "dodge", binwidth = 0.05) + 
facet_wrap(~generation) +
guides(alpha = FALSE) +
labs(title = "Suicide rates across generations", subtitle = "Countries have high suicide rate for slice", x = "Suicide rate (in %)")

```

## Trend of suicide rate over time 
```{r}
global_avg_suicide_no <- (sum(as.numeric(master$suicides_no)) / sum(as.numeric(master$population))) * 100000
p0 <- master %>%
  group_by(year) %>%
  summarize(population = sum(population), 
            suicides = sum(suicides_no), 
            suicides_per_100k = (suicides / population) * 100000) %>%
  ggplot(aes(x = year, y = suicides_per_100k)) + 
  geom_line(col = "#9932CC", size = 1) + 
  geom_point(col = "#9932CC", size = 2) + 
  geom_hline(yintercept = global_avg_suicide_no, linetype = 2, color = "#D8BFD8", size = 1) + 
  geom_text(aes(x = 2011, y=global_avg_suicide_no, label = "Global Avg Suicides Line", vjust = -1)) +
  labs(title = "Trend of Global Suicides per 100k Over time, 1985 - 2015",
       subtitle = "Suicides number has decreased significantly since 1995",
       x = "Year", 
       y = "Suicides per 100k") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5)) +
   theme(axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        panel.background = element_blank()) 
p0
ggsave('trend.png', p0)
```

```{r warnning= FALSE}
p1 <- master %>%
  group_by(age, sex) %>%
  summarize(suicides = sum(suicides_no)) %>%
  ggplot(aes(x = age, y = suicides, fill = reorder(sex, suicides))) + 
  geom_bar(stat = "identity") +  
  labs(title = "Total Suicides From Each Age Group By Gender",
       subtitle = "Men at age between 35-54 may under the most pressure") +
  scale_fill_manual(values = c('grey', '#FF99FF'),
                    name = '',
                    labels = c('Female', "Male"))+
  scale_x_discrete(limit = c("5-14", "15-24", "35-54","55-74","75+")) +
  scale_y_continuous(labels = scales::comma)+
   theme(axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        panel.background = element_blank(),
        legend.direction = "horizontal",
        legend.position = "top") 

p1
ggsave('age.png', p1)
```

## World map with suicide rate

```{r}
# grey area indicates missing data
pdf("map.pdf",width=6,height=4,paper='special') 

country <- master %>%
  group_by(country) %>%
  summarize(country_suicide_no = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

countrydata <- joinCountryData2Map(country, joinCode = "NAME", nameJoinColumn = "country")

par(mar=c(0, 0, 0, 0))

mapCountryData(countrydata, 
               nameColumnToPlot="country_suicide_no", 
               mapTitle="", 
               colourPalette = "heat", 
               oceanCol="lightblue", 
               missingCountryCol="grey65", 
               catMethod = "pretty")

dev.off()
```



# EDA Summary

+ European countries have significantly higher suicide rates
+ Suicide rate reaches to peach from 1991 to 2003, where 1995 has the highest 



# Statistical analysis

Since the data has the same count from each categorical vairable, there is no need to conduct statistical test (like chi-square test) on these categorical variables. So I am going to conduct the regression model directly. 

## Multiple linear regression

Null hypothesis: suicides number has no relationships with sex, age, GDP per capita and continent

```{r}
# remove units and brand
ma <- master[, -c(1,7,9,10)]

mod <-  lm(suicides_no ~ sex + age  + gdp_per_capita + continent, data = ma)

#summary(mod)

# Pull out the coefficients and confidence interval for table and graph
coe <- summary(mod)$coefficients # get coefficients and related stats
coe_CI <- as.data.frame(cbind(coe[-1, ], confint(mod)[-1, ])) # find and bind CI, remove Intercept

# Rename results data frame
names(coe_CI) <- c("estimate", "se", "t", "pval","low_CI","high_CI")

htmlTable(round(coe_CI[order(coe_CI$pval, decreasing = FALSE), ],3))

```

## Model diagnosis

```{r}
par(mfrow = c(2,3))
plot(ma$year,mod$residuals)
plot(ma$sex,mod$residuals)
plot(ma$age,mod$residuals)
plot(ma$population,mod$residuals)
plot(ma$gdp_per_capita,mod$residuals)
plot(ma$continent,mod$residuals)

```


```{r}
model_diag <- plot_model(mod, type ="diag")
model_diag[1]

```

```{r}
model_diag[3]
```

```{r}
model_diag[4]
```

## Statistical analysis Summary

+ Sex, age, continent are statistically and practically significant
    - Except for independ variables, like continent$Oceania and age (55-74 group)
+ suicide number is expected to increase 0.02 for every dollar increase in GDP per capita, holding others constant.

# Save Image of file
```{r}
# # Save the rds file to reuse anything from this file in another file
# save.image("suicide.RData")
# 
# # recover using load()
# # load("office_TA.RData")
```

